---
title: "Product Choice Logit"
author: "Amelia Morrissey"
date: "2/11/2021"
output: html_document
---

```{r}
library(readxl)
library(dplyr)
library(lubridate)
library(tidyr)
library(tibble)
library(data.table)
```

#Read in the Filtered Data
```{r}
Filtered <- read_excel("C:/Users/staff/PVTA Dropbox/PVTA-PVPC/Mobile Data Analysis/Data/Formatted Data/Archive/Filtered_03312021.xlsx")

Filtered <- Filtered_03312021
```

Filter for UUID accounts that only use the four major fare types exclusively. 
This includes filtering out E&D riders for now.
-PPU (One Ride, Transfers)
-Day Passes
-7 Day Passes
-31 Day Passes

Format
```{r}
#mark any UUIDs who have used non-main pass types, and rename pass Faretypes
#to combine E&D and Regular since that distinction will be on the account
#level rather than the pass level
Product_Choice <- Filtered %>%
  group_by(UUID) %>%
  mutate(filter_out = any(Faretype == "Children's One Ride " | 
         Faretype == "Senior Tuesday " |
         Faretype == "Holyoke Community College Fall Semester " |
         Faretype == "Holyoke Community College Spring Semester ")) %>%
  filter(!filter_out) %>%
  mutate(Faretype = ifelse(Faretype == "One Ride E & D ", "One Ride ", Faretype)) %>%
  mutate(Faretype = ifelse(Faretype == "Transfer E & D ", "Transfer", Faretype)) %>%
  mutate(Faretype = ifelse(Faretype == "31 Day E & D ", 
                        "31 Day Pass ", Faretype)) %>%
  mutate(Faretype = ifelse(Faretype == "31 Day Regular ", 
                        "31 Day Pass ", Faretype)) 

#Also change PPU passes to be called PPU
Product_Choice <- Product_Choice %>% 
   mutate(Faretype = ifelse(Faretype == "One Ride ", "PPU", Faretype)) %>%
   mutate(Faretype = ifelse(Faretype == "Transfer", "PPU", Faretype))

#remove the first week of data
Product_Choice <- Product_Choice %>% 
  filter(Date >= ymd("2020-07-27"))


#separate E&D riders from non-E&D
#we will consider these groups separately
# Product_Choice_Reg <- Product_Choice %>%
#   filter(!ED)
# 
Product_Choice_ED <- Product_Choice %>%
  filter(ED)
```

Aggregate on Day Level and Get User's Choice
```{r}
Product_Choice_Reg <- Product_Choice
#aggregate on the daily Level
Product_Choice_Reg <- Product_Choice_Reg %>%
  mutate(count = 1) %>%
  group_by(Date, UUID) %>%
  summarise(D31_Rides = sum(count[Faretype == "31 Day Pass "]),
            D7_Rides = sum(count[Faretype == "7 Day Pass"]),
            Day_Rides = sum(count[Faretype == "Day Pass"]),
            PPU_Rides = sum(count[Faretype == "PPU"]) )
Product_Choice_Reg$rowname <- as.character(seq.int(nrow(Product_Choice_Reg)))

# Get the user's "choice" on that day, i.e. the most used pass type
#library(tidyr)
#library(tibble)
temp <- Product_Choice_Reg %>% 
  ungroup() %>%
  select(D31_Rides, D7_Rides, Day_Rides, PPU_Rides) 
temp <- temp %>%
  rownames_to_column() %>%
  gather(column, value, -rowname) %>%
  group_by(rowname) %>% 
  filter(rank(-value) == 1) %>%
  select(-value)

Product_Choice_Reg <- left_join(Product_Choice_Reg, temp)

#for any cases where the person used an equal amount of two pass types
#we will just remove these cases since we don't know their choice
Product_Choice_Reg <- Product_Choice_Reg %>%
  dplyr::rename(choice = column) %>%
  filter(!is.na(choice))
```

Find Use Value and Active Days from the previous week
```{r}

# #remove the first week since it's ride count will not include a full 7 days
# Product_Choice <- Product_Choice %>%
#   filter(Date >= lubridate::ymd("2020-07-27"))

#count the number of rides within the past 7 days

#for each UUID Date in Product_Choice_Reg, create an interval for the past 7 days
#then intersect






#this takes up a lot of memory so break the data set into sections by UUID
Product_Choice <- Product_Choice %>%
  mutate(UUID_section = cut(UUID, breaks = seq(0, 3000, by = 100)))

Product_Choice$UUID_section <- Product_Choice %>%
  group_indices(UUID_section)

sections = n_distinct(Product_Choice$UUID_section)

#initialize week rides by running through with i = 1
i = 1 
data <- Product_Choice %>% 
  ungroup() %>%
  filter(UUID_section == i) %>%
  select(UUID, Date) %>%
  mutate(Cust_ID = as.character(UUID),
         Record_Date = as.Date(Date)) %>%
  select(-UUID, -Date)
data <- as.data.table(data)

data_1 <- distinct(data)
  

week_rides <- 
  data[,as.list(data[,list(Cust_ID2=Cust_ID,Compare_Date=Record_Date)]),
  by=c("Cust_ID","Record_Date")][Cust_ID==Cust_ID2 & Compare_Date<Record_Date,list(Cust_ID,Record_Date,Compare_Date,
    Within7=(as.numeric(Record_Date-Compare_Date)<=7)*1)][,list(Within7=(sum(Within7))),by=c("Cust_ID","Record_Date")]  


week_active <-
  data_1[,as.list(data_1[,list(Cust_ID2=Cust_ID,Compare_Date=Record_Date)]),
  by=c("Cust_ID","Record_Date")][Cust_ID==Cust_ID2 & Compare_Date<Record_Date,list(Cust_ID,Record_Date,Compare_Date,
    Within7=(as.numeric(Record_Date-Compare_Date)<=7)*1)][,list(Within7=(sum(Within7))),by=c("Cust_ID","Record_Date")]  


for(i in (2:sections)){
data <- Product_Choice %>% 
  ungroup() %>%
  filter(UUID_section == i) %>%
  select(UUID, Date) %>%
  mutate(Cust_ID = as.character(UUID), 
         Record_Date = as.Date(Date)) %>%
  select(-UUID, -Date)
data <- as.data.table(data)

data_1 <- distinct(data)

output_rides <- 
  data[,as.list(data[,list(Cust_ID2=Cust_ID,Compare_Date=Record_Date)]),
  by=c("Cust_ID","Record_Date")][Cust_ID==Cust_ID2 & Compare_Date<Record_Date,list(Cust_ID,Record_Date,Compare_Date,
    Within7=(as.numeric(Record_Date-Compare_Date)<=7)*1)][,list(Within7=(sum(Within7))),by=c("Cust_ID","Record_Date")]      

output_active <-
  data_1[,as.list(data_1[,list(Cust_ID2=Cust_ID,Compare_Date=Record_Date)]),
  by=c("Cust_ID","Record_Date")][Cust_ID==Cust_ID2 & Compare_Date<Record_Date,list(Cust_ID,Record_Date,Compare_Date,
    Within7=(as.numeric(Record_Date-Compare_Date)<=7)*1)][,list(Within7=(sum(Within7))),by=c("Cust_ID","Record_Date")]  

#add these rides to the week_rides and week_active dfs
week_rides = rbind(week_rides, output_rides)
week_active = rbind(week_active, output_active)
}

#reformat for joining
week_rides <- week_rides %>%
  dplyr::rename("UUID" = "Cust_ID",
                "Date" = "Record_Date") %>%
  dplyr::rename("week_rides" = "Within7") %>%
  mutate(UUID = as.integer(UUID))
  
week_active <- week_active %>%
  dplyr::rename("UUID" = "Cust_ID",
                "Date" = "Record_Date") %>%
  dplyr::rename("week_active" = "Within7") %>%
  mutate(UUID = as.integer(UUID))
  
```

Join in the past week's rides and active days
```{r}
#join the week_rides  and week_active days to the Product_Choice df
Product_Choice_Reg <- left_join(Product_Choice_Reg, week_rides)
Product_Choice_Reg <- left_join(Product_Choice_Reg, week_active)



```

```{r}
#replace past week cost and choice for passholding days with 
#the pass start day's choice scenario

#get first and last day of each pass ID for non-PPU
temp <- Product_Choice %>%
  filter(Faretype != "PPU" & Faretype != "Day Pass") %>%
  group_by(Id) %>%
  mutate(min_date = min(date(FTime_Scanned))) #%>%
  # mutate(First = ifelse((date(FTime_Scanned) == min_date), 
  #                       T, F))#,
         #last_date = ifelse(Faretype == "7 Day Pass", min_date+7, min_date+31))

#if a date is a first passholding day, it remains the same
#but if it is not first and it is a pass Faretype choice, then replace the 
#choice scenario with the most recent first day by that UUID
temp <- temp %>%
  ungroup() %>%
  select(UUID, Date, min_date) %>%
  distinct()

Product_Choice_Reg <- left_join(Product_Choice_Reg, temp)

temp1 <- Product_Choice_Reg %>%
  filter(min_date == Date) %>%
  ungroup() %>%
  select(min_date, UUID, week_rides, week_active) %>%
  dplyr::rename("week_rides2" = "week_rides",
                "week_active2" = "week_active")

Product_Choice_Reg1 <- left_join(temp1, Product_Choice_Reg)
Product_Choice_Reg <- left_join(Product_Choice_Reg, Product_Choice_Reg1)

#fill in the ppu and day pass week rides and active days
Product_Choice_Reg <- Product_Choice_Reg %>%
  mutate(week_rides2 = ifelse(is.na(week_rides2) & is.na(min_date), 
                              week_rides, week_rides2)) %>%
  mutate(week_active2 = ifelse(is.na(week_active2) & is.na(min_date), week_active, week_active2))

rm(Product_Choice_Reg1)


```

Do we need to fill in non-active pass-holding days with the initial pass purchase choice scenario?

Remove choices after 7+ days of inactivity
```{r}
#NA indicates that this is the first day this user has ever been active
#we will remove these, as well as any dates with 0 rides in the past week 
#This includes any passholding days where the original pass choice scenario was 
#the first in at least 7 days, so we use week_rides2 for the passes
Product_Choice_Reg <- Product_Choice_Reg %>%
  filter(!is.na(week_rides2) & week_rides2 != 0)

table(Product_Choice_Reg$choice)
```

Calculate past week's cost for each fare type
```{r}
# One_Ride_Fare = 1.5
One_Ride_Fare = 0.75
# Transfer_Fare = 0.25
Transfer_Fare = 0.1
Day_Fare = 3.5
D7_Fare = 15
# D31_Fare = 54
D31_Fare = 26



Product_Choice_Reg <- Product_Choice_Reg %>%
  mutate(week_cost_PPU = round(
           (week_rides2-week_rides2*evasion_trans_rate)*One_Ride_Fare +
           week_rides2*evasion_trans_rate*Transfer_Fare, digits=2),
         week_cost_Day = Day_Fare*week_active2,
         week_cost_7D = D7_Fare, 
         week_cost_31D = round(D31_Fare*(7/31), digits=2))

```

Remove unnecessary variables
```{r}
Product_Choice_Reg <- Product_Choice_Reg %>%
  select(-rowname, -D31_Rides, -D7_Rides, -Day_Rides, -PPU_Rides, 
         -week_rides, -week_active, -min_date) %>%
  dplyr::rename("week_rides" = "week_rides2",
                "week_active" = "week_active2")
```

Bin the choice scenarios by the past weeks use value
```{r}
# Bin_Width = 4
# Top_Bin = 32 + Bin_Width
# 
# 
# #break into use value bins
# Product_Choice_Reg <- Product_Choice_Reg %>%
#   mutate(week_cost_PPU_capped = ifelse(week_cost_PPU > Top_Bin, 
#                                       Top_Bin, week_cost_PPU)) %>%
#   mutate(week_cost_Bin = cut(week_cost_PPU_capped, breaks = 
#                                seq(0, Top_Bin, by = Bin_Width)))
```


Calculate Weights to Scale up to entire ridership population from mobile
```{r}
#create FY21 Actuals df
Ridership_Actuals <- data.frame(choice = c("PPU", "Day", 
                                         "7D", "31D"),
                 FY21Jul_Dec_Rides = c(552845, 290410, 45722, 321636) )

#including E&D rides: c(552845, 290410, 45722, 321636) )


#calculate weights
Weights <- Product_Choice_Reg %>%
  group_by(choice) %>%
  summarise(choice_days = n_distinct(UUID, Date)) %>%
 #  dplyr::rename("Pass" = "choice") #%>%
  mutate(choice = ifelse(choice == "PPU_Rides", "PPU", choice)) %>%
  mutate(choice = ifelse(choice == "Day_Rides", "Day", choice)) %>%
  mutate(choice = ifelse(choice == "D7_Rides", "7D", choice)) %>%
  mutate(choice = ifelse(choice == "D31_Rides", "31D", choice))

Weights <- left_join(Weights, Ridership_Actuals)

Weights <- Weights %>%
  mutate(weight = FY21Jul_Dec_Rides/choice_days) %>%
  mutate(weight = round(weight, digits = 2)) %>%
  select(choice, weight)

#wide format#
##################

# #create FY21 Actuals df
# Ridership_Actuals <- data.frame(choice.PPU = c(1,0,0,0),
#                                 choice.Day = c(0,1,0,0),
#                                 choice.D7 = c(0,0,1,0),
#                                 choice.D31 = c(0,0,0,1),
#                  FY21Jul_Dec_Rides = c(552845, 290410, 45722, 321636) )

#calculate weights
Weights <- Product_Choice_Reg %>%
  group_by(choice) %>%
  summarise(choice_days = n_distinct(UUID, Date)) %>%
 #  dplyr::rename("Pass" = "choice") #%>%
  mutate(choice = ifelse(choice == "PPU_Rides", "PPU", choice)) %>%
  mutate(choice = ifelse(choice == "Day_Rides", "Day", choice)) %>%
  mutate(choice = ifelse(choice == "D7_Rides", "7D", choice)) %>%
  mutate(choice = ifelse(choice == "D31_Rides", "31D", choice))

Weights <- left_join(Weights, Ridership_Actuals)

Weights <- Weights %>%
  mutate(weight = FY21Jul_Dec_Rides/choice_days) %>%
  mutate(weight = round(weight, digits = 2)) %>%
  select(choice, weight)
```



Calculate weights based on use value bin
```{r}
# #get the total number of trips in each bin
# Bin_Trips <- Filtered %>%
#   ungroup() %>%
#   mutate(filter_out = (Faretype == "Children's One Ride " | 
#          Faretype == "Senior Tuesday " |
#          Faretype == "Holyoke Community College Fall Semester " |
#          Faretype == "Holyoke Community College Spring Semester ")) %>%
#   filter(!filter_out) %>%
#   mutate(Faretype = ifelse(Faretype == "One Ride E & D ", "One Ride ", Faretype)) %>%
#   mutate(Faretype = ifelse(Faretype == "Transfer E & D ", "Transfer", Faretype)) %>%
#   mutate(Faretype = ifelse(Faretype == "31 Day E & D ", 
#                         "31 Day Pass", Faretype)) %>%
#   mutate(Faretype = ifelse(Faretype == "31 Day Regular ", 
#                         "31 Day Pass", Faretype)) %>%
#   select(-filter_out)
# 
# #Also change PPU passes to be called PPU
# Bin_Trips <- Bin_Trips %>% 
#    mutate(Faretype = ifelse(Faretype == "One Ride ", "PPU", Faretype)) %>%
#    mutate(Faretype = ifelse(Faretype == "Transfer", "PPU", Faretype))
# 
# Bin_Trips <- Bin_Trips %>%
# filter(!ED) %>%
#   group_by(Use_Value_evasion_Bin, Faretype) %>%
#   summarise(trips = n()) %>%
#   dplyr::rename("week_cost_Bin" = "Use_Value_evasion_Bin")
# 
# 
# #get number of days in the choice days data bins
# Bin_Days <- Product_Choice_Reg %>%
#   group_by(week_cost_Bin, choice) %>%
#   summarise(choice_days = n())
# 
# #calculate weights
# Bin_Weights <- left_join(Bin_Days, Bin_Trips)
# Bin_Weights <- Bin_Weights %>%
#   mutate(weight = round(trips/choice_days, digits=2))
```



Data Expansion (Long format) and Mlogit regression
```{r}
mydata <- Product_Choice_Reg %>%
  ungroup() %>%
  select(choice, week_cost_PPU, week_cost_Day, week_cost_7D, week_cost_31D) %>%
  dplyr::rename("week_cost.PPU" = "week_cost_PPU",
                "week_cost.Day" = "week_cost_Day",
                "week_cost.7D" = "week_cost_7D",
                "week_cost.31D" = "week_cost_31D")

mydata <- mydata %>%
  mutate(choice = ifelse(choice == "PPU_Rides", "PPU", choice),
         choice = ifelse(choice == "Day_Rides", "Day", choice),
         choice = ifelse(choice == "D7_Rides", "7D", choice),
         choice = ifelse(choice == "D31_Rides", "31D", choice)
         )

# mydata <- mydata %>%
#   mutate(choice.PPU = ifelse(choice == "PPU_Rides", 1, 0),
#          choice.Day = ifelse(choice == "Day_Rides", 1, 0),
#          choice.7D = ifelse(choice == "D7_Rides", 1, 0),
#          choice.31D = ifelse(choice == "D31_Rides", 1, 0)
#          )

#add weights to mydata
library(mlogit)
mydata <- left_join(mydata, Weights)
mydata$choice <- as.factor(mydata$choice)
mldata <- dfidx(mydata, varying=2:5, choice="choice", shape="wide")

mlogit <- mlogit(choice ~ week_cost, data=mldata, reflevel="PPU")

summary(mlogit)

```

```{r}
#For Day Passes,what is the averge num. trips per day?
Day <- Filtered %>%
  filter(Faretype == "Day Pass")

Day<- Day %>%
  group_by(UUID, Date) %>%
  summarise(n = n())

mean(Day$n)
```


